---
title: "Used code to obtain results"
output: pdf_document
---

# 1. Execution of models in WinBUGS using the library R2WinBUGS

```{r echo=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, tidy = TRUE, tidy.opts = list(blank = FALSE, width.cutoff = 40))
```

## Load libraries and data
```{r}
# Working directory
DirMain = " " # Set an appropriate directory
setwd(DirMain)
# Load library and data
library(R2WinBUGS)
library(knitr)
library(pbugs) # For running the models in parallel calls to WinBUGS
load("datos.Rdata")
```

## BYM model

```{r}
# BYM model with independent diseases, WinBUGS code
BYM.indep <- function() {
  for (j in 1:Ndiseases) {
    for(i in 1:Nareas) {
    O[i, j] ~ dpois(lambda[i, j])
    # Modeling of the mean for each municipality
    log(lambda[i, j]) <- log(E[i, j]) + mu[j] + sdhet[j] * het[i, j] + sdsp[j] * sp[j, i]
    # SMR for each municipality
    SMR[i] <- exp(mu[j] + sdhet[j] * het[i, j] + sdsp[j] * sp[j, i])
    # Prior distribution for the non-spatial effect
    het[i, j] ~ dnorm(0, 1)
  }
  # Prior distribution for the spatial effect
  sp[j, 1:n] ~ car.normal(adj[], w[], num[], 1)
  # Prior distribution for the mean risk for all municipalities
  mu[j] ~ dflat()
  # Prior distribution for the standard deviations of the random effects	
  sdhet[j] ~ dunif(0, 5)
  sdsp[j] ~ dunif(0, 5)
  }
}

# Run BYM model for each city, sex and disease

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j <- 2 
    # Sex = 1: Mens, 2: Women
    for(k in 1:2) { 
      # Mortality cause
      l <- c(1:20)
        
# Matrix with observed and expected cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
E <- t(apply(Esp[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
Nareas <- dim(O)[1]
Ndiseases <- dim(O)[2]

  # Data        
  data <- list(Nareas = Nareas, Ndiseases = Ndiseases, O = O, E = E, adj = unlist(nb[[i]]), 
               weights = rep(1, length(unlist(nb[[i]]))), num = sapply(nb[[i]], length))
  # Initial values
  initials <- function(){list(mu = rnorm(Ndiseases, 0, 0.1), 
              sdhet = runif(Ndiseases, 0, 1), sdsp = runif(Ndiseases, 0, 1), 
              het = matrix(rnorm(Nareas * Ndiseases), nrow = Ndiseases, ncol = Nareas), 
              sp = matrix( rnorm( Nareas * Ndiseases), nrow = Ndiseases, ncol = Nareas))}
  # Variables to retrive
  param <- c("SMR", "lambda", "het", "sp", "mu", "sdsp", "sdhet")
  # Calls to WinBUGS        
  t.result <- system.time(result <- pbugs(data = data, model.file = BYM.indep, 
                                          inits = initials, parameters.to.save = param, 
                                          n.chains = 3, n.iter = 30000, n.burnin = 5000, 
                                          DIC = F))
  # Save results        
  save(result, t.result, file = paste(getwd(), "/resul/resul.BYMIndep.", i,".", j,".", 
                                      k, ".Rdata", sep=""))
      }
}
```

## Fixed effects $\bf M$-model (Section 2 in paper)

```{r}
# Fixed effects M-model, WinBUGS code

Mmodel.FE <- function() {
  for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			O[i, j] ~ dpois(lambda[i, j])
      # Modeling of the mean for each municipality and disease
			log(lambda[i, j]) <- log(E[i, j]) + mu[j] + Theta[i, j]
      # SMR for each municipality and disease
      SMR[i, j] <- exp(mu[j] + Theta[i, j])
		}
	}

# Definition of the random effects matrix
	for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			Theta[i, j] <- inprod2(tPhi[, i], M[, j])
		}
	}

# Matrix of spatially correlated random effects
# if M is a square matrix define Nsp (Number of spatial underlying patterns) as Ndiseases
	for(j in 1:Nsp) {
# Prior distribution for the spatial effect	
		Spatial[j, 1:Nareas] ~ car.normal(adj[], weights[], num[], 1)
		for(i in 1:Nareas) {
# Prior distribution for the non-spatial effect
			Het[j, i] ~ dnorm(0, 1)
			tPhi[j, i] <- Spatial[j, i]
		}
	} 
  
	for(j in (Nsp + 1):(2 * Nsp)) {
		for(i in 1:Nareas) {
			tPhi[j, i] <- Het[(j - Nsp), i]
		}
	}	
	
# M-matrix
	for(i in 1:(2 * Nsp)) {
		for(j in 1:Ndiseases) {
      M[i, j] ~ dflat()
		}
	}		

# Others prior distributions
	for(j in 1:Ndiseases) {
# Prior distribution for the mean risk for all municipalities
		mu[j] ~ dflat()
	}
}

# Run fixed effects M-model considering 20 causes of mortality

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3){ 
  # Specific mortality causes
  j<-2 
    # Sex = 1: Mens, 2: Women
    for(k in 1:2){ 
      # Mortality causes 
      l<-c(1:20)  

# Matrix with observed and expected cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
E <- t(apply(Esp[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
Nareas <- dim(0)[1]
Ndiseases <- dim(O)[2]
# Data
data <- list(Nareas = Nareas, Ndiseases = Ndiseases, Nsp = Ndiseases, O = O, E = E, adj = unlist(nb[[i]]), weights = rep(1, length(unlist(nb[[i]]))), num = sapply(nb[[i]], length))
# Initial values
initials <- function(){list(mu = rnorm(Ndiseases, 0, 0.1), M = matrix(rnorm(2 * Nsp * Ndiseases), nrow = 2 * Nsp, ncol = Ndiseases), Het = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas), Spatial = matrix( rnorm( Nareas * Nsp), nrow = Nsp, ncol = Nareas))}
# Variables to retrieve
param <- c("SMR", "lambda", "M", "Het", "Spatial", "mu", "Theta")
# Calls to WinBUGS
t.result <- system.time(result <- pbugs(data = data, model.file = Mmodel.FE, inits = initials, parameters.to.save = param, n.chains = 3, n.iter = 30000, n.burnin = 5000, DIC = F))
# Save results
save(result, t.result, file = paste(getwd(), "/resul/resul.MmodelFE.", i,".", j,".", k, ".Rdata", sep=""))

    }
}
```

## Random effects $\bf M$-model or NVA $\mathbf M$-model (Section 2 in paper)

```{r}
# Random effects M-model, WinBUGS code

Mmodel.RE <- function() {
  for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			O[i, j] ~ dpois(lambda[i, j])
# Modeling of the mean for each municipality and disease
			log(lambda[i, j]) <- log(E[i, j]) + mu[j] + Theta[i, j]
# SMR for each municipality and disease
      SMR[i, j] <- exp(mu[j] + Theta[i, j])
		}
	}

# Definition of the random effects matrix
	for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			Theta[i, j] <- inprod2(tPhi[, i], M[, j])
		}
	}

# Matrix of spatially correlated random effects:
# if M is a square matrix define Nsp (Number of spatial underlying patterns) as Ndiseases
	for(j in 1:Nsp) {
# Prior distribution for the spatial effect	
		Spatial[j, 1:Nareas] ~ car.normal(adj[], weights[], num[], 1)
		for(i in 1:Nareas) {
# Prior distribution for the non-spatial effect
			Het[j, i] ~ dnorm(0, 1)
			tPhi[j, i] <- Spatial[j, i]
		}
	} 
  
	for(j in (Nsp + 1):(2 * Nsp)) {
		for(i in 1:Nareas) {
			tPhi[j, i] <- Het[(j - Nsp), i]
		}
	}	
	
# M-matrix
  for(j in 1:Ndiseases){
    for(i in 1:Nsp){
      M[i, j] ~ dnorm(0, prec.sp)
    }
    
    for(i in (Nsp+1):(2*Nsp)){
      M[i, j] ~ dnorm(0, prec.het)
    }
  }

# Others prior distributions
	
# Prior distribution for the mean risk for all municipalities
	for(j in 1:Ndiseases) {
		mu[j] ~ dflat()
	}
# Prior distribution for the standard deviations of the random effects		
  prec.sp <- pow(sd.sp, -2)
	sd.sp ~ dunif(0, 100)
	 
	prec.het <- pow(sd.het, -2)
	sd.het ~ dunif(0, 100)
}

# Run random effects M-model considering 20 causes of mortality

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j<-2 
    # Sex = 1: Mens, 2: Women
    for(k in 1:2) { 
      # Mortality causes 
      l<-c(1:20)  

# Matrix with observed and expected cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
E <- t(apply(Esp[[i]][[j]][k, , , ], c(1,2), sum)[l, ])

Nareas <- dim(0)[1]
Ndiseases <- dim(O)[2]

# Data
data <- list(Nareas = Nareas, Ndiseases = Ndiseases, Nsp = Ndiseases, O = O, E = E, adj = unlist(nb[[i]]), weights = rep(1, length(unlist(nb[[i]]))), num = sapply(nb[[i]], length))

# Initial values
initials <- function(){list(mu = rnorm(Ndiseases, 0, 0.1), sd.sp = runif(1, 0, 1), sd.het = runif(1, 0, 1), Het = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas), Spatial = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas))}

# Variables to retrieve
param <- c("sd.sp", "sd.het", "SMR", "lambda", "M", "Het", "Spatial", "mu")

# Calls to WinBUGS
t.result <- system.time(result <- pbugs(data = data, model.file = Mmodel.RE, inits = initials, parameters.to.save = param, n.chains = 3, n.iter = 30000, n.burnin = 5000, DIC = F))

# Save results
save(result, t.result, file = paste(getwd(), "/resul/resul.MmodelRE.", i,".", j,".", k, ".Rdata", sep=""))
    }
}
  
```

## RVA $\mathbf M$-model (Section 4 in paper)

```{r}
# RVA M-model, WinBUGS code

Mmodel.RVA <- function() {
  for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			O[i, j] ~ dpois(lambda[i, j])
# Modeling of the mean for each municipality and disease
			log(lambda[i, j]) <- log(E[i, j]) + mu[j] + Theta[i, j]
# SMR for each municipality and disease
      SMR[i, j] <- exp(mu[j] + Theta[i, j])
		}
	}

# Definition of the random effects matrix
	for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			Theta[i, j] <- inprod2(tPhi[, i], M[, j])
		}
	}

# Matrix of spatially correlated random effects
# if M is a square matrix define Nsp (Number of spatial underlying patterns) as Ndiseases
	for(j in 1:Nsp) {
# Prior distribution for the spatial effect	
		Spatial[j, 1:Nareas] ~ car.normal(adj[], weights[], num[], 1)
		for(i in 1:Nareas) {
# Prior distribution for the non-spatial effect
			Het[j, i] ~ dnorm(0, 1)
			tPhi[j, i] <- Spatial[j, i]
		}
	} 
  
	for(j in (Nsp + 1):(2 * Nsp)) {
		for(i in 1:Nareas) {
			tPhi[j, i] <- Het[(j - Nsp), i]
		}
	}	
	

# M-matrix
  for(j in 1:Ndiseases) {

      for(i in 1:(2 * Nsp)) {
        M.aux[i, j] ~ dnorm(0, 1)      
        M[i, j] <- sd[i] * M.aux[i, j]    
      }
  }

# Others prior distributions

# Prior distribution for the mean risk for all municipalities 
	for(j in 1:Ndiseases) {
		mu[j] ~ dflat()
	}
  
# Prior distribution for the standard deviations of the random effects		
	for(i in 1:(2 * Nsp)) {
    sd[i] ~ dunif(0, 5)
  }
}

# Run RVA M-model considering 20 causes of mortality 

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j<-2 
    # Sex = 1: Mens, 2: Women
    for(k in 1:2) { 
      # Mortality causes
      l<-c(1:20)   

# Matrix with observed and expected cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
E <- t(apply(Esp[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])

Nareas <- dim(0)[1]
Ndiseases <- dim(O)[2]

# Data
data <- list(Nareas = Nareas, Ndiseases = Ndiseases, Nsp = Ndiseases, O = O, E = E, adj = unlist(nb[[i]]), weights = rep(1, length(unlist(nb[[i]]))), num = sapply(nb[[i]], length))

# Initial values
initials <- function(){list(mu = rnorm(Ndiseases, 0, 0.1), sd = runif(2 * Nsp, 0.1, 1), Het = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas), Spatial = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas))}

# Variables to retrieve
param <- c("sd", "SMR", "lambda", "M", "mu")

# Calls to WinBUGS
t.result <- system.time(result <- pbugs(data = data, model.file = Mmodel.RVA, inits = initials, parameters.to.save = param, n.chains = 3, n.iter = 30000, n.burnin = 5000, DIC = F))

# Save results
save(result, t.result, file = paste(getwd(), "/resul/resul.MmodelRVA.", i, ".", j, ".", k, ".Rdata", sep=""))
    }
}
```

## CVA $\mathbf M$-model (Section 4 in paper)

```{r}
# CVA M-model, WinBUGS code

Mmodel.CVA <- function() {
  for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			O[i, j] ~ dpois(lambda[i, j])
# Modeling of the mean for each municipality and disease
			log(lambda[i, j]) <- log(E[i, j]) + mu[j] + Theta[i, j]
# SMR for each municipality and disease
      SMR[i, j] <- exp(mu[j] + Theta[i, j])
		}
	}

# Definition of the random effects matrix
	for(i in 1:Nareas) {
		for(j in 1:Ndiseases) {
			Theta[i, j] <- inprod2(tPhi[, i], M[, j])
		}
	}

# Matrix of spatially correlated random effects
# if M is a square matrix define Nsp (Number of spatial underlying patterns) as Ndiseases
	for(j in 1:Nsp) {
# Prior distribution for the spatial effect	
		Spatial[j, 1:Nareas] ~ car.normal(adj[], weights[], num[], 1)
		for (i in 1:Nareas) {
		# Prior distribution for the non-spatial effect
			Het[j, i] ~ dnorm(0, 1)
			tPhi[j, i] <- Spatial[j, i]
		}
	} 
  
	for(j in (Nsp + 1):(2 * Nsp)){
		for(i in 1:Nareas) {
			tPhi[j, i] <- Het[(j - Nsp), i]
		}
	}	

# M-matrix
  for(j in 1:Ndiseases) {
      for(i in 1:Nsp) {
        M.aux[i, j] ~ dnorm(0, 1)      
        M[i,j] <- sdstruct.sp[j] * M.aux[i, j]
      }
      for(i in (Nsp + 1):(2 * Nsp)){
        M.aux[i, j] ~ dnorm(0, 1)      
        M[i, j] <- sdstruct.het[j] * M.aux[i, j]
      }
  }

# Others prior distributions
# Prior distribution for the mean risk for all municipalities 
	for(j in 1:Ndiseases) {
		mu[j] ~ dflat()
	}
  
# Prior distribution for the standard deviations of the random effects		
		for(j in 1:Ndiseases) {
      prec.sp[j] <- pow(sdstruct.sp[j], -2)
      sdstruct.sp[j] ~ dunif(0, 5)

      prec.het[j] <- pow(sdstruct.het[j], -2)
      sdstruct.het[j] ~ dunif(0,5)
  }
}

# Run CVA M-model considering 20 causes of mortality 

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j<-2 
    # Sex = 1: Mens, 2: Women
    for(k in 1:2){ 
      # Mortality causes 
      l<-c(1:20)  

# Matrix with observed and expected cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])
E <- t(apply(Esp[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])

Nareas <- dim(0)[1]
Ndiseases <- dim(O)[2]

# Data
data <- list(Nareas = Nareas, Ndiseases = Ndiseases, Nsp = Ndiseases, O = O, E = E, adj = unlist(nb[[i]]), weights = rep(1, length(unlist(nb[[i]]))), num = sapply(nb[[i]], length))

# Initial values
initials <- function(){list(mu = rnorm(Ndiseases, 0, 0.1), sdstruct.sp = runif(Ndiseases, 0, 1), sdstruct.het = runif(Ndiseases, 0, 1),Het = matrix(rnorm(Nareas * Nsp), nrow = Nsp, ncol = Nareas), Spatial = matrix(rnorm(Nareas * Nsp), nrow=Nsp, ncol = Nareas))}

# Variables to retrieve
param <- c("sdstruct.sp", "sdstruct.het", "SMR", "lambda", "M", "mu")

# Calls to WinBUGS
t.result <- system.time(result <- pbugs(data = data, model.file = Mmodel.CVA, inits = initials, parameters.to.save = param, n.chains = 3, n.iter = 30000, n.burnin = 5000, DIC = F))

# Save results
save(result, t.result, file = paste(getwd(), "/resul/resul.MmodelCVA.", i, ".", j, ".", k, ".Rdata", sep=""))
    }
}
```

# 2. Choropleth maps for all models

```{r}
# Load libraries
library(RColorBrewer)

cuts_SMR <- c(0, 0.67, 0.80, 0.91, 1.10, 1.25, 1.50)
palette <- brewer.pal(7, "BrBG")[7:1]

# Name of mortality causes 
Causes <- dimnames(Obs[[i]][[j]])[[2]]

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j<-2 
  # Sex = 1: Mens, 2: Women
    k<-1 
      # Mortality causes 
      for(l in 1:20) {  
      
# Load WinBUGS results, BYM model
load(paste(getwd(), "/resul.BYMIndep.", i, ".", j, ".", k, ".", l, ".Rdata", sep = ""))

# SMR estimates, BYM model
aux <- palette[findInterval(result$mean$SMR, cuts_SMR)]
plot(Carto[[i]], col = palette[aux], main = paste0("BYM - ", Causas[l]), lwd = 0.2)
legend("bottomright", c("< 0.67","0.67 - 0.80","0.80 - 0.91","0.91 - 1.10","1.10 - 1.25","1.25 - 1.50","> 1.50"), title = "SMR", border = NULL, fill = palette, bty = "n")

# Load WinBUGS results, fixed effects M-model
load(paste(getwd(), "/resul.MmodelFE." , i, ".", j, ".", k,".Rdata", sep = ""))

# SMR estimates, fixed effects M-model
aux <- palette[findInterval(result$mean$SMR[, l], cuts_SMR)]
plot(Carto[[i]], col = palette[aux], main = paste0("MmodelFE - ", Causas[l]), lwd = 0.2)
legend("bottomright", c("< 0.67", "0.67 - 0.80", "0.80 - 0.91", "0.91 - 1.10", "1.10 - 1.25", "1.25 - 1.50", "> 1.50"), title="SMR", border=NULL, fill = palette, bty = "n")

# Load WinBUGS results, random effects M-model
load(paste(getwd(), "/resul.MmodelRE.", i, ".", j, ".", k, ".Rdata", sep=""))

# SMR estimates, random effects M-model
aux <- palette[findInterval(result$mean$SMR[, l], cuts_SMR)]
plot(Carto[[i]], col = palette[aux], main = paste0("MmodelRE - ", Causas[l]), lwd = 0.2)
legend("bottomright", c("< 0.67", "0.67 - 0.80", "0.80 - 0.91", "0.91 - 1.10", "1.10 - 1.25", "1.25 - 1.50", "> 1.50"), title = "SMR", border = NULL, fill = palette, bty = "n")

# Load WinBUGS results, RVA m-model
load(paste(getwd(), "/resul.MmodelRVA.", i, ".", j, ".", k, ".Rdata", sep = ""))

# SMR estimates, RVA M-model
aux <- palette[findInterval(result$mean$SMR[, l], cuts_SMR)]
plot(Carto[[i]], col = palette[aux], main = paste0("MmodelRVA - ", Causas[l]), lwd = 0.2)
legend("bottomright", c("< 0.67", "0.67 - 0.80", "0.80 - 0.91", "0.91 - 1.10", "1.10 - 1.25", "1.25 - 1.50", "> 1.50"), title = "SMR", border = NULL, fill = palette, bty = "n")

# Load WinBUGS results, CVA m-model
load(paste(getwd(), "/resul.MmodelCVA.", i, ".", j, ".", k, ".Rdata", sep = ""))

# SMR estimates, CVA M-model
aux <- palette[findInterval(result$mean$SMR[, l], cuts_SMR)]
plot(Carto[[i]], col = palette[aux], main = paste0("MmodelCVA - ", Causas[l]), lwd = 0.2)
legend("bottomright", c("< 0.67", "0.67 - 0.80", "0.80 - 0.91", "0.91 - 1.10", "1.10 - 1.25", "1.25 - 1.50", "> 1.50"), title = "SMR", border = NULL, fill = palette, bty = "n")

      }
}
```

# 3. DIC for each model (Section 5, Table 1 in paper)

```{r}
 
# Function for DICs calculation
CalculaDIC <- function(Simu, O, save = FALSE) {
  mu <- Simu$sims.matrix[, which(substr(dimnames(Simu$sims.matrix)[[2]], 1, 2) == "la")]
  D <- apply(mu, 1, function(x){ -2 * sum(dpois(as.vector(t(O)), x, log = T))})
  Dmedia <- mean(D)
  mumedia <- apply(mu, 2, mean)
  DenMedia <- -2 * sum(dpois(as.vector(t(O)), mumedia, log = T))
  if(save == TRUE){ return(c(Dmedia, Dmedia - DenMedia, 2 * Dmedia - DenMedia))}
  cat("D=", Dmedia, "pD=", Dmedia - DenMedia, "DIC=", 2 * Dmedia - DenMedia, "\n")
}

DIC.BYMIndep <- list()
DIC.MmodelFE <- list()
DIC.MmodelRE <- list()
DIC.MmodelRVA <- list()
DIC.MmodelCVA <- list()

# City = 1: Alicante, 2: Castellón, 3: Valencia
for(i in 1:3) { 
  # Specific mortality causes
  j <- 2 
    # Sex = 1: Mens, 2: Women
    k <- 1 
      # Mortality causes 
      l <- c(1:20)  

# Matrix with observed cases
O <- t(apply(Obs[[i]][[j]][k, , , ], c(1, 2), sum)[l, ])

# DIC M-model with independent diseases 
load(paste(getwd(), "/resul.BYMIndep.", i, ".", j, ".", k, ".Rdata", sep = ""))
DIC.BYMIndep[[i]] <- CalculaDIC(Simu = result, O = O, save = TRUE)[3]
# DIC fixed effects M-model
load(paste(getwd(), "/resul.MmodelFE.", i, ".", j, ".", k, ".Rdata", sep = ""))
DIC.MmodelFE[[i]] <- CalculaDIC(Simu = result, O = O, save = TRUE)[3]
# DIC random effects M-model
load(paste(getwd(), "/resul.MmodelRE.", i,".", j,".", k, ".Rdata", sep = ""))
DIC.MmodelRE[[i]] <- CalculaDIC(Simu = result, O = O, save = TRUE)[3]
# DIC RVA M-model
load(paste(getwd(), "/resul.MmodelRVA.", i, ".", j, ".", k, ".Rdata", sep = ""))
DIC.MmodelRVA[[i]] <- CalculaDIC(Simu = result, O = O, save = TRUE)[3]
# DIC CVA M-model
load(paste(getwd(), "/resul.MmodelCVA.", i, ".", j, ".", k, ".Rdata", sep = ""))
DIC.MmodelCVA[[i]] <- CalculaDIC(Simu = result, O = O, save = TRUE)[3]

    }

Table <- matrix(c(unlist(DIC.BYMIndep), unlist(DIC.MmodelFE), unlist(DIC.MmodelRE), unlist(DIC.MmodelRVA), unlist(DIC.MmodelCVA)), ncol = 3, byrow = TRUE)
rownames(Table) <- c("BYM with independent diseases", "Fixed effects $M$-model", "Random effects $M$-model", "RVA $M$-model", "CVA $M$-model")
colnames(Table) <- c("Alicante", "Castellón", "Valencia")
print(kable(Table, caption = "DICs for the adjusted models in each study city"))
```

# 4. Log-likelihood of the predictive distribution of each model in the cases observed in odd years (Section 5, Table 2 in paper)

```{r}

log_predict.BYMIndep <- array(dim = c(3, length(l), result$n.sims))
log_predict.MmodelFE <- array(dim = c(3, length(l), result$n.sims))
log_predict.MmodelRE <- array(dim = c(3, length(l), result$n.sims))
log_predict.MmodelRVA <- array(dim = c(3, length(l), result$n.sims))
log_predict.MmodelCVA <- array(dim = c(3, length(l), result$n.sims))

# City = 1: Alicante, 2: Castellón, 3: Valencia
for (i in 1:3) { 
  # Specific mortality causes
  j <- 2 
  # Sex = 1: Mens, 2: Women
  k <- 1
  # Mortality causes 
  l <- c(1:20)
    
  # Matrix with observed and expected cases in odd years
  O.odd <- t(apply(Obs[[i]][[j]][k, , , seq(2, 12, by = 2)], c(1, 2), sum)[l, ])
  E.odd <- t(apply(Esp[[i]][[j]][k, , , seq(2, 12, by = 2)], c(1, 2), sum)[l, ])
  
  # Load results M-model with independent diseases (couple years)
  load(paste(getwd(), "/resul.BYMIndep.couple.", i, ".", j, ".", k, ".Rdata", sep=""))
  # Load results fixed effects M-model (couple years)
  load(paste(getwd(), "/resul.MmodelFE.couple.", i, ".", j, ".", k, ".Rdata", sep=""))
  # Load results random effects M-model (couple years)
  load(paste(getwd(), "/resul.MmodelRE.couple.", i, ".", j, ".", k, ".Rdata", sep=""))
  # Load results RVA M-model (couple years)
  load(paste(getwd(), "/resul.MmodelRVA.couple.", i, ".", j, ".", k, ".Rdata", sep=""))
  # Load results CVA M-model (couple years)
  load(paste(getwd(), "/resul.MmodelCVA.couple.", i, ".", j, ".", k, ".Rdata", sep=""))
      
  for(m in 1:length(l)){
    for(n in 1:result$n.sims){
      
      # M-model with independent diseases 
      lambda <- E.odd[, m] * resul.BYMIndep.couple$sims.list$SMR[n, , m]
      log_predict.BYMIndep[i, m, n] <- sum(dpois(O.odd[, m], lambda, log = TRUE))
      # Fixed effects M-model
      lambda <- E.odd[, m] * resul.MmodelFE.couple$sims.list$SMR[n, , m]
      log_predict.MmodelFE[i, m, n] <- sum(dpois(O.odd[, m], lambda, log = TRUE))
      # Random effects M-model
      lambda <- E.odd[, m] * resul.MmodelRE.couple$sims.list$SMR[n, , m]
      log_predict.MmodelRE[i, m, n] <- sum(dpois(O.odd[, m], lambda, log = TRUE))
      # RVA M-model
      lambda <- E.odd[, m] * resul.MmodelRVA.couple$sims.list$SMR[n, , m]
      log_predict.MmodelRVA[i, m, n] <- sum(dpois(O.odd[, m], lambda, log = TRUE))
      # CVA M-model
      lambda <- E.odd[, m] * resul.MmodelCVA.couple$sims.list$SMR[n, , m]
      log_predict.MmodelCVA[i, m, n] <- sum(dpois(O.odd[, m], lambda, log = TRUE))
    }
  }
}

Table <- matrix(c(apply(log_predict.BYMIndep, 1, mean), apply(log_predict.MmodelFE, 1, mean), apply(log_predict.MmodelRE, 1, mean), apply(log_predict.MmodelRVA, 1, mean), apply(log_predict.MmodelCVA, 1, mean)), ncol = 3, byrow = TRUE)

rownames(Table) <- c("BYM with independent diseases", "Fixed effects $M$-model", "Random effects $M$-model", "RVA $M$-model", "CVA $M$-model")
colnames(Table) <- c("Alicante", "Castellón", "Valencia")
print(kable(Table, caption = "Log-likelihood of the predictive distribution of each model 
            in the cases observed in odd years and in each study city"))

```

